MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-26 main/MSstatsPTM_Main.tex
Small

We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.  Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$. In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$ \todo{check the use of hat}. The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in \eqref{msstats_model}. The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$. Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$, producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$.

From the summaries above, the proposed approach estimates the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp}
\begin{eqnarray}
\hat{\Delta}_{adj} = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i}) - (\hat{\mu}_{i^{\prime}}^{\ast}-\hat{\mu}_{i^{\prime}}) = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}) - (\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}) = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein} 
\label{eq:Delta_hat_adj}
\end{eqnarray}
Assuming that the sources of variation in the modified features unexplained by the model are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits
\begin{eqnarray}
\widehat{SE}(\hat{\Delta}_{adj}) = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
\label{eq:prop_se}
\end{eqnarray}
For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error would be calculated as 
\begin{eqnarray}
 \widehat{SE}(\hat{\Delta}_{adj}) &=& \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}}
\label{eq:se_calc}
\end{eqnarray}
The estimated standard error is larger than the standard errors associated with each distinct feature type, reflecting the increased uncertainty inherent to combining the uncertainty in two estimators.

Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{kutner_etal_04a, satterthwaite:1946}
\begin{eqnarray}
df \left(  \widehat{SE}(\hat{\Delta}_{adj}) \right)=  \left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
\left( \frac{\widehat{SE}(\hat{\Delta}_{PTM})^{4}}{\mathrm{df}(\widehat{SE}(\hat{\Delta}_{PTM}))} + \frac{\widehat{SE}(\hat{\Delta}_{protein})^{4}}{ \mathrm{df}(\widehat{SE}(\hat{\Delta}_{protein}))} \right)
\label{eq:prop_df}
\end{eqnarray}
To test the null hypothesis in~\eqref{null_hyp}, the test statistic $\hat{\Delta}_{adj}/\widehat{SE}(\Delta)_{adj}$ is compared with the Student distribution with the degrees of freedom in~\eqref{prop_df}. The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.

\todo{This paragraph relates more to the implementation. Could you integrate it with the implementation text?} By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile. It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs. It is applicable to label-free data acquisition strategies such as DDA and DIA and experiments with TMT labeling. The proposed approach is also applicable to experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.

\subsubsection*{Design of balanced PTM experiments with one source of variation}
\label{sec:design}

The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively. The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations. With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}. Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta}_{adj})$ as

\begin{equation}
\widehat{SE}(\hat{\Delta}_{adj})^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{PTM}^{2} + \hat{\sigma}_{protein}^{2} \right) \right]
\leq \left( \frac{\Delta_{adj}}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
\end{equation}
where 
\begin{equation}
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1}
\end{equation}

and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs.

Solving for $J$, the number of biological replicates per group is calculated as

\begin{equation}
J \geq \frac{(2\hat{\sigma}_{PTM}^{2} + 2\hat{\sigma}_{protein}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta_{adj}^2}
\end{equation}

Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in {\bf Supplementary Sec. \ref{supp-sec:complex_design}}. More details on sample size calculation can be found in~\cite{oberg_vitek_09a}.

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}. $MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT. $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut. We require the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e. the modification site). This will generally include the amino acid abbreviation, plus it's number in the protein sequence. For example, a modification on a serine, which was the 70th amino acid in the sequence would be marked ``S70''.  
Occasionally tools will only include the peptide sequence with the modified amino acid highlighted. If there is no way to determine where in the protein sequence the modification occurred, $MSstatsPTM$ includes functionality to do this manually, given the modified peptide sequence and a FASTA file with the entire protein sequence. The package includes converters for seamless integration with multiple such tools. The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites. If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect. Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest. For experimental designs that include separate runs enriched in modified peptides and global profiling runs, conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.

>>>>

We leverage the existing methods in $MSstats$ and $MSstatsTMT$, and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.  Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$. In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$ \todo{check the use of hat}. The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in \eqref{msstats_model}. The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$. Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, and the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$ producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$.

From the summaries above, the proposed approach estimates the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp}
\begin{eqnarray}
\hat{\Delta}_{adj} = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i}) - (\hat{\mu}_{i^{\prime}}^{\ast}-\hat{\mu}_{i^{\prime}}) = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}) - (\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}) = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein} 
\label{eq:Delta_hat_adj}
\end{eqnarray}
Assuming that the sources of variation in the modified features that are unexplained by the model are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits
\begin{eqnarray}
\widehat{SE}(\hat{\Delta}_{adj}) = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
\label{eq:prop_se}
\end{eqnarray}
For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are respectively  the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error is calculated as 
\begin{eqnarray}
 \widehat{SE}(\hat{\Delta}_{adj}) &=& \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}}
\label{eq:se_calc}
\end{eqnarray}
The estimated standard error is larger than the standard errors associated with each individual feature type, reflecting the combined uncertainty in the two estimates. 
Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{kutner_etal_04a, satterthwaite:1946}
\begin{eqnarray}
df \left(  \widehat{SE}(\hat{\Delta}_{adj}) \right)=  \left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
\left( \frac{\widehat{SE}(\hat{\Delta}_{PTM})^{4}}{\mathrm{df}(\widehat{SE}(\hat{\Delta}_{PTM}))} + \frac{\widehat{SE}(\hat{\Delta}_{protein})^{4}}{ \mathrm{df}(\widehat{SE}(\hat{\Delta}_{protein}))} \right)
\label{eq:prop_df}
\end{eqnarray}
To test the null hypothesis in~\eqref{null_hyp}, the test statistic $\hat{\Delta}_{adj}/\widehat{SE}(\Delta)_{adj}$ is compared with the Student distribution with the degrees of freedom in~\eqref{prop_df}. The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.

\todo{This paragraph relates more to the implementation. Could you integrate it with the implementation section?} By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile. It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs. It is applicable to label-free data acquisition strategies such as DDA and DIA and experiments with TMT labeling. The proposed approach is also applicable to experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.

\subsubsection*{Sample size calculation for future PTM experiments}
\label{sec:design}

The proposed statistical framework enables sample size calculation for future PTM experiments. 
The procedure has been described in general in~\cite{kutner_etal_04a}, and specifically for protein significance analysis in~\cite{oberg_vitek_09a}. 
It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the summaries of the modified and unmodified features, respectively. Typically, the variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow. Given the above quantities, and assuming a balanced design, the minimal number of replicates $J$ across $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\hat{\Delta}_{adj})$: \todo{There should be no hats in equations 11-13. The calculations assume that the variances are given externally, and are known. The input quantities above have no hats}
\begin{equation}
\widehat{SE}(\hat{\Delta}_{adj})^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{PTM}^{2} + \hat{\sigma}_{protein}^{2} \right) \right]
\leq \left( \frac{\Delta_{adj}}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
\end{equation}
where 
\begin{equation}
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1}
\end{equation}
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Student distribution with $df = I(J-1)$ degrees of freedom.  Solving for $J$, the number of biological replicates per condition is
\begin{equation}
J \geq \frac{(2\hat{\sigma}_{PTM}^{2} + 2\hat{\sigma}_{protein}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta_{adj}^2}
\end{equation}
\todo{Comment on the fact that since the numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.} Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec. \ref{supp-sec:complex_design}}.

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}. $MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT. $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.

$MSstatsPTM$ requires the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e. the modification site). This will generally include the amino acid abbreviation, plus it's number in the protein sequence. For example, a modification on a serine, which was the 70th amino acid in the sequence would be marked ``S70''.

Occasionally tools will only include the peptide sequence with the modified amino acid highlighted. If there is no way to determine where in the protein sequence the modification occurred, $MSstatsPTM$ includes functionality to do this manually, given the modified peptide sequence and a FASTA file with the entire protein sequence. The package includes converters for seamless integration with multiple such tools. The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites. If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect. Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest. For experimental designs that include separate runs enriched in modified peptides and global profiling runs, conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.
